Revisiting the hydrodynamic modulation of short surface waves by longer waves

### 2025-03-08 paper.tex
Addressing minor reviews

\begin{abstract}
Hydrodynamic modulation of short ocean surface waves by longer ambient waves is
a well-known ocean surface process that affects remote sensing, the
interpretation of \textit{in situ} wave measurements, and numerical wave
forecasting.
In this paper, we revisit the wave crest and action conservation laws and derive
higher-order steady solutions for the change of short-wave wavenumber, action,
and gravitational acceleration due to the presence of longer waves.
We validate the analytical solutions with numerical simulations of the full wave
crest and action conservation equations.
The nonlinear analytical solutions of short-wave wavenumber, amplitude, and
steepness modulation significantly deviate from the linear analytical solutions
of \citet{longuet1960changes}, and are similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution}.
The short-wave steepness modulation is attributed to be 5/8 due to the
wavenumber, 1/4 due to the wave action, and 1/8 due to the effective gravity.
We further examine the results of \citet{peureux2021unsteady} who found through
numerical simulations that the short-wave modulation grows unsteadily with
each long-wave passage.
We show that this unsteady growth only occurs for homogeneous initial
conditions as a special case and does not generally occur, for example in more
realistic long-wave groups.
The proposed steady solutions are a good approximation of the fully
nonlinear numerical solutions in long-wave steepness up to $\approx$0.2.
Except for a subset of initial conditions, the solutions to the fully nonlinear
crest-action conservation equations are mostly steady in the reference frame of
the long waves.
\end{abstract}

@@

\begin{equation}
\label{eq:lhs1960wavenumber}
\widetilde{k} = k (1 + \varepsilon_L \cos{\psi})
\end{equation}
where $\widetilde{k}$ and $k$ are the modulated and unmodulated short-wave wavenumbers,
respectively, and $\varepsilon_L = a_L k_L$ is the steepness of the long wave with
wavenumber $k_L$ and amplitude $a_L$, and $\psi$ is the long-wave phase.
This process is illustrated in Fig. \ref{fig:hydrodynamic_modulation_diagram}.
Hydrodynamic modulation is a key process for the development and
interpretation of remote sensing of ocean surface waves and currents
\citep{keller1975microwave,hara1994hydrodynamic}, and is distinguished
from aerodynamic modulation \citep{donelan1987effect,belcher1999wave,chen2000effects}.
\citet{phillips1981dispersion} and \citet{longuet1987propagation} revisited the
hydrodynamic modulation problem by considering nonlinear long waves and
variations in the effective gravity of short waves.
Their results yielded significantly stronger modulation than previously predicted.
\citet{henyey1988energy} derived an analytical solution for the modulation of
short waves using Hamiltonian mechanics and reported similar modulation magnitudes
to that of \citet{longuet1987propagation}.
\citet{zhang1990evolution} considered weakly nonlinear short waves
using a nonlinear Schr\" odinger equation and found stronger wavenumber
modulation but weaker amplitude modulation than those predicted by
\citet{longuet1987propagation}.
The variety of analytical and numerical frameworks (Eulerian crest and action
conservation, Hamiltonian, Schr\" odinger) that tackle the same fundamental
physics and that relatively closely reproduce steady solutions for short-wave
modulation suggests that the solutions are robust.
Nevertheless, alternative analytical and numerical approaches to the modulation
problem remain of interest, especially if higher degrees of nonlinearity can be
considered using simpler approaches.

@@

In this paper, we revisit this problem from the perspective of linear wave theory
but consider the long-wave slope to be significant enough to require evaluating
the short-waves at the long-wave surface rather than at the mean water level.
This allows us to derive alternative, steady, nonlinear solutions for the
modulation of short waves, within the limits of small $\varepsilon_L$.
The analytical solutions yield modulations that are stronger than the original
solution by \citet{longuet1960changes}, but similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution} in
long-wave steepness up to $\approx 0.2$.
To validate the steadiness of the analytical solutions, we perform numerical
simulations of the full wave crest and action conservation laws, and compare
the results to the analytical solutions presented here, as well as the prior
numerical results.
The numerical simulations are used to determine the validity of the approximate
steady solution and to investigate the unsteadiness of the short-wave modulation
found by \citet{peureux2021unsteady}.
The main prior studies on hydrodynamic modulation, and their assumptions and
approaches to the solution, are summarised in Table
\ref{table:modulation_literature_summary}.
We first present the governing equations in \S\ref{section:governing_equations},
and then the analytical solutions for the modulation of short waves by long waves
in \S\ref{section:analytical_solutions}.
Then, we describe the numerical simulations and their results in
\S\ref{section:numerical_solutions}.
Finally, we discuss the results and their implications in
\S\ref{section:discussion} and conclude the paper in \S\ref{section:conclusions}.

@@

\begin{equation}
\label{eq:k_short_lhs}
\frac{\widetilde{k}}{k} = 1 + \varepsilon_L \cos{\psi}
\end{equation}
Coming from the conservation of crests, as opposed to the velocity potential
superposition, the solution by \citet{longuet1960changes} requires two
approximations:
First, evaluating the long-wave velocity at $z = 0$ rather than $z = \eta_L$; and
second, further truncating the Taylor expansion series beyond the first-order
term.
These approximations underestimate the short-wave modulation magnitude
(Fig. \ref{fig:analytical_solutions_ak0.1}a).
At the wave crest, (\ref{eq:k_short_exact}) yields the wavenumber modulation
that is 16.9\%, 38.5\%, 66.4\%, and 104\% larger than that of
\citet{longuet1960changes} for $\varepsilon_L = 0.1, 0.2, 0.3, 0.4$, respectively.
Although not immediately obvious from the expressions, the modulated wavenumber
(\ref{eq:k_short_lhs}) is conserved across the long-wave phase, whereas
the higher-order solution (\ref{eq:k_short_exact}) is not.
This is because combining the orbital velocities evaluated at $z=\eta_L$ and the
linearised conservation of wave crests in (\ref{eq:wavenumber}) creates excess
short-wave wavenumber.
The solution of (\ref{eq:wavenumber}) must be conservative across the long-wave
phase, as we will see later from the numerical simulations.
Although the analytical solutions by \citet{longuet1960changes} has been updated
by the numerical simulations of \citet{longuet1987propagation}, they are here
nevertheless relevant to use as reference for the proposed analytical solutions.

@@

\begin{equation}
\label{eq:gravity_modulation_lagrangian}
\widetilde{g_l} =
g + \frac{dW}{dt} =
g + \frac{\partial W}{\partial t} + \left(C_g + U\right) \frac{\partial W}{\partial x} =
\widetilde{g_e} + \left(C_g + U\right) \frac{\partial W}{\partial x}
\end{equation}
where $l$ and $e$ subscripts serve to denote "Lagrangian" and "Eulerian",
respectively.
Although \citet{longuet1986eulerian} thoroughly described the differences
between the Eulerian and Lagrangian effective gravities, he did so for
a fully nonlinear irrotational gravity wave, which can be computed numerically
but is not readily available analytically.
Note also that when evaluating the Lagrangian effective gravity,
\citet{longuet1987propagation} considered the orbital motion of the long wave
but not the propagation speed of the short-wave group, which can exceed the
ambient velocity $U$ depending on the properties of the short and long waves
considered.
Group speed of the short waves aside, the Lagrangian correction to the gravity
being an advective term is $\mathcal{O}(\varepsilon_L^2)$.
Third, the projection of the long-wave induced horizontal acceleration
contributes small but non-negligible effects on the short-wave effective
gravity in the curvilinear coordinate system
\citep{phillips1981dispersion,zhang1990evolution}.
These three effects are completely described by projecting the gravitational
acceleration vector from the coordinate that is perpendicular to the mean water
surface ($z=0$) to that which is perpendicular to the long-wave surface
($z=\eta_L$), and likewise for the long-wave orbital accelerations:

@@

\begin{equation}
\label{eq:gravity_modulation_linear}
\frac{\widetilde{g_e}}{g}
  = 
  \frac{
    1 - \varepsilon_L \cos{\psi} e^{\varepsilon_L \cos{\psi}}
    \left[ 1 + \left(\varepsilon_L \sin{\psi}\right)^2 \right]
  }
  {\sqrt{\left(\varepsilon_L \sin{\psi}\right)^2 + 1}}
\end{equation}
The same derivation can be carried out for the Lagrangian gravity
modulation following (\ref{eq:gravity_modulation_lagrangian})-(\ref{eq:local_slope}),
and is omitted here for brevity.
Curvilinear effects on effective gravity can be removed altogether by setting
$\alpha = 0$.
In that case, (\ref{eq:gravity_modulation_general}) simplifies to:

\begin{equation}
\label{eq:gravity_modulation_linear_no_curvature}
\frac{\widetilde{g}}{g} = 1 - \varepsilon_L \cos{\psi} e^{\varepsilon_L \cos{\psi}}
\end{equation}
If we evaluate the orbital accelerations at the mean water level $z=0$,
this further simplifies to:

\begin{equation}
\label{eq:gravity_modulation_linear_no_curvature_mean_level}
\frac{\widetilde{g}}{g} = 1 - \varepsilon_L \cos{\psi}
\end{equation}
which is the form used by \citet{peureux2021unsteady}, for example.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/fig_effective_gravities.pdf}
\caption{
  Analytical solutions for the effective gravitational acceleration
  modulation by long waves, as function of the long-wave phase, for $\varepsilon_L = 0.3$.
  Black is for Eulerian gravity of a linear wave evaluated at $z=0$;
  blue is the same as black but evaluated at $z=\eta$;
  orange is the same as blue but for Lagrangian gravity;
  green is the same as orange but in a curvilinear reference frame;
  red is the same as green but for a third-order Stokes wave;
  and purple is the same as red but for a fully-nonlinear wave.
  The elevation and surface velocities for the fully-nonlinear wave are computed
  following \citet{clamond2018accurate}.
}
\label{fig:effective_gravities}
\end{figure}

@@

To quantify the contribution of the nonlinear terms in (\ref{eq:wavenumber})
and (\ref{eq:wave_action2}), and to evaluate the steadiness assumption of the
analytical solutions, we now proceed to numerically integrate the full set of
crest and action conservation equations.
(\ref{eq:wavenumber}) and (\ref{eq:wave_action2}) are discretised using
second-order central finite difference in space and integrated in time using the
fourth-order Runge-Kutta method \citep{butcher1996runge}.
The space is divided into 100 grid points.
This is effectively the same equation set and numerical configuration as that of
\citet{peureux2021unsteady}, except that their long-wave orbital velocities are
evaluated at $z=0$ instead of $z=\eta_L$, thus neglecting the Stokes drift
induced by long waves on short-wave groups
\citep{stokes1847,van2018stokes,monismith2020stokes}.
Their gravity modulation also does not consider the nonlinear or the curvature
effects, however, this difference does not qualitatively affect the results.
Another difference is the choice of the numerical scheme for spatial differences,
which in their case is the more sophisticated MUSCL4 scheme \citep{kurganov2000new}.
Although the second-order central finite difference is not appropriate for many
numerical problems, in our case it is sufficient because it is conservative and
the fields that we compute derivatives of are smooth and continuous.
The maximum relative error of the centred finite difference in this model
is $\approx 0.066\%$.
The wavenumber and wave action are conservative over time within
$\mathcal{O}(10^{-7})$ and $\mathcal{O}(10^{-5})$ relative error, respectively.
We show in the next section that our numerical solutions for infinite long-wave
trains are qualitatively equivalent to those of \citet{peureux2021unsteady}.

The numerical equations here are integrated in a fixed reference frame,
rather than that moving with the long-wave phase speed.
Either approach produces equivalent modulation results, however a fixed
reference frame may allow for a more intuitive interpretation of the results.
This difference is important to keep in mind when comparing the results
of \citet{peureux2021unsteady} and those presented here; in their Fig. 1 the
long-wave phase is fixed and the short waves are moving leftward
with the speed of $C_{pL} - C_g - U$ (neglecting group speed inhomogeneity),
whereas in the figures that follow, the long wave is moving rightward with its
phase speed $C_{pL}$ and the short waves (that is, their action) are moving
rightward with the speed of $C_g + U$ (neglecting group speed inhomogeneity).
The crest and action conservation equations are integrated in the curvilinear
coordinate system in which the horizontal coordinate follows the long-wave
surface, akin to \citet{zhang1990evolution}.

@@

The first set of simulations is performed for the long-wave steepness of
$\varepsilon_L = 0.1$, $0.2$, and $0.4$ (Fig. \ref{fig:numerical_solutions}).
The long-waves are initialised from rest ($a_L = 0$) and gradually ramped up
to their target steepness over 5 long-wave periods, thus emulating the passage
of a long-wave group
(the importance of which we discuss in more detail in the next subsection).
At $\varepsilon_L = 0.1$, the numerical solutions for the wavenumber, amplitude,
and steepness modulation are all remarkably similar to the analytical solutions
(Fig. \ref{fig:numerical_solutions}a,d,g).
This suggests that for low long-wave steepness the steady analytical solution
is a reasonable proxy for the fully-nonlinear crest-action conservation
equation solutions.
The numerical solutions diverge more notably from the analytical ones at
moderate steepness of 0.2 (Fig. \ref{fig:numerical_solutions}b,e,h), and are
considerably different at very high steepness of 0.4
(Fig. \ref{fig:numerical_solutions}c,f,i).
These differences suggest that at high steepness the inhomogeneity of
gravitational acceleration and group speed, otherwise ignored in the steady
solutions, become important.

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/numerical_vs_analytical_solutions_k.pdf}
\caption{
  Maximum wavenumber modulation as function of long-wave steepness $\varepsilon_L$.
  Black line and circles are for the analytical and numerical solutions from
  \citet{longuet1960changes} and \citet{longuet1987propagation}, respectively.
  Blue line is based on the analytical solutions from this paper.
  Green and orange lines are for the numerical solutions from this paper
  using linear and third-order Stokes long waves, respectively.
}
\label{fig:numerical_vs_analytical_k}
\end{figure}

@@

The maximum modulations of the short-wave wavenumber, amplitude, and steepness
from the analytical and numerical solutions, are shown on Figs.
\ref{fig:numerical_vs_analytical_k},
\ref{fig:numerical_vs_analytical_a}, and
\ref{fig:numerical_vs_analytical_ak}, respectively, as function of the long-wave
steepness $\varepsilon_L$.
Overall, the numerical solutions, based on the linear and the third-order Stokes
long waves alike, begin to diverge from the analytical solution at
$\varepsilon_L \approx 0.2$.
The numerical solution using the Stokes long wave is quantitatively very similar
to the fully-nonlinear solutions of \citet{longuet1987propagation} in amplitude
modulation, however with somewhat lower wavenumber and steepness modulations.
For example, at $\varepsilon_L = 0.4$, the steepness modulation in our solutions
is close to 6, whereas \citet{longuet1987propagation} finds it close to 8.
The underestimate in the wavenumber modulation at high long-wave steepness
relative to the results of \citet{longuet1987propagation}
is likely due to his long wave velocities being evaluated to a higher order in
$\varepsilon_L$ than they are here.
Nevertheless, the overall steepness modulation between the two numerical
approaches are remarkably similar, and either solution would cause the
short-waves of even low steepness to break \citep{banner1993wave}.
Where the two modulation solutions differ (at $\varepsilon_L > 0.2$),
either solution would lead to short-wave breaking after one long-wave period.
This suggests that in the cases where the short-waves would survive the
modulation and persist over multiple long-wave periods, the two solutions are
effectively equivalent.

As the numerical solutions are for nonlinear wave action and crest conservation
equations, it is important to re-evaluate the homogeneity and stationarity
requirements for the wave action conservation to remain valid.
Fig. \ref{fig:homogeneity_stationarity_numerical} shows the homogeneities
and stationarities of the short-wave wavenumber, action, and effective gravity
as function of the long-wave steepness $\varepsilon_L$ and the wavenumber ratio
$k/k_L$, based on the numerical simulations.
Although homogeneity is slightly lower ($H_{k,N,g} \approx 0.92$ for $k/k_L = 10$
and $\varepsilon_L = 0.4$) compared to the analytical solutions,
the stationarity is significantly reduced.
Following the same criteria for strong ($S_{k,N,g} > 0.99$) and weak
($S_{k,N,g} > 0.9$) stationarity as in subsection \ref{sec:wave_action_conservation},
the numerical solutions are only weakly stationary for $\varepsilon_L < 0.21$
at $k/k_L = 10$, and for $\varepsilon_L < 0.36$ at $k/k_L = 100$.
This provides a quantitative guidance on when the short-wave action can be
considered to be conserved.

>>>>

\begin{abstract}
Hydrodynamic modulation of short ocean surface waves by longer ambient waves is
a well-known ocean surface process that affects remote sensing, the
interpretation of \textit{in situ} wave measurements, and numerical wave
forecasting.
In this paper, we revisit the wave crest and action conservation laws and derive
higher-order steady solutions for the change of short-wave wavenumber, action,
and gravitational acceleration due to the presence of longer waves.
We validate the analytical solutions with numerical simulations of the full wave
crest and action conservation equations.
The nonlinear analytical solutions of short-wave wavenumber, amplitude, and
steepness modulation significantly deviate from the linear analytical solutions
of \citet{longuet1960changes}, and are similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution}.
The short-wave steepness modulation is attributed to be 5/8 due to the
wavenumber, 1/4 due to the wave action, and 1/8 due to the effective gravity.
Examining the homogeneity and stationarity requirements of the wave action
conservation equation reveals that stationarity is more easily violated than
homogeneity, and that the short-wave action is strongly conserved ($>$99\%) only
for very small amplitude long waves, and weakly conserved ($>$90\%) for steep
long waves.
We further examine the results of \citet{peureux2021unsteady} who found through
numerical simulations that the short-wave modulation grows unsteadily with
each long-wave passage.
We show that this unsteady growth only occurs for homogeneous initial
conditions as a special case and does not generally occur, for example in more
realistic long-wave groups.
The proposed steady solutions are a good approximation of the fully
nonlinear numerical solutions in long-wave steepness up to $\approx$0.2.
Except for a subset of initial conditions, the solutions to the fully nonlinear
crest-action conservation equations are mostly steady in the reference frame of
the long waves.
\end{abstract}

@@

\begin{equation}
\label{eq:lhs1960wavenumber}
\widetilde{k} = k (1 + \varepsilon_L \cos{\psi})
\end{equation}
where $\widetilde{k}$ and $k$ are the modulated and unmodulated short-wave wavenumbers,
respectively, and $\varepsilon_L = a_L k_L$ is the steepness of the long wave with
wavenumber $k_L$ and amplitude $a_L$, and $\psi$ is the long-wave phase.
This process is illustrated in Fig. \ref{fig:hydrodynamic_modulation_diagram}.
Hydrodynamic modulation is a key process for the development and
interpretation of remote sensing of ocean surface waves and currents
\citep{keller1975microwave,alpers1978two,hara1994hydrodynamic}, and is distinguished
from aerodynamic modulation \citep{donelan1987effect,belcher1999wave,chen2000effects}.
Experimental studies of the general short-wave modulation problem included both
the laboratory \citep[e.g.,][]{keller1975microwave,donelan2010modulation} and field
\citep[e.g.,][]{hara2003observation,plant1977studies} measurements.
\citet{phillips1981dispersion} and \citet{longuet1987propagation} revisited the
hydrodynamic modulation problem by considering nonlinear long waves and
variations in the effective gravity of short waves.
Their results yielded significantly stronger modulation than previously predicted
by \citet{longuet1960changes}.
\citet{henyey1988energy} derived an analytical solution for the modulation of
short waves using Hamiltonian mechanics and reported similar modulation magnitudes
to that of \citet{longuet1987propagation}.
\citet{zhang1990evolution} considered weakly nonlinear short waves
using a nonlinear Schr\" odinger equation and found stronger wavenumber
modulation but weaker amplitude modulation than those predicted by
\citet{longuet1987propagation}.
Aside from the early linear solutions by \citet{longuet1960changes}, the variety
of analytical and numerical frameworks (Eulerian crest and action conservation,
Hamiltonian, Schr\" odinger) that tackle the same fundamental
physics and that relatively closely reproduce the steady solutions for short-wave
modulation suggests that the solutions are robust.
Nevertheless, alternative analytical and numerical approaches to the modulation
problem remain of interest, especially if higher degrees of nonlinearity can be
considered using simpler approaches.

@@

In this paper, we revisit this problem from the perspective of wave action
conservation but consider the long-wave slope to be significant enough to
require evaluating the short-waves at the long-wave surface rather than at the
mean water level.
This allows us to derive alternative, steady, nonlinear solutions for the
modulation of short waves, within the limits of small $\varepsilon_L$.
The analytical solutions yield modulations that are stronger than the original
solution by \citet{longuet1960changes}, but similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution} in
long-wave steepness up to $\approx 0.2$.
To validate the steadiness of the analytical solutions, we perform numerical
simulations of the full wave crest and action conservation laws, and compare
the results to the analytical solutions presented here, as well as the prior
numerical results.
The numerical simulations are used to determine the validity of the approximate
steady solution and to investigate the unsteadiness of the short-wave modulation
found by \citet{peureux2021unsteady}.
The main prior studies on hydrodynamic modulation, and their assumptions and
approaches to the solution, are summarised in Table
\ref{table:modulation_literature_summary}.
We first present the governing equations in \S\ref{section:governing_equations},
and then the analytical solutions for the modulation of short waves by long waves
in \S\ref{section:analytical_solutions}.
Then, we describe the numerical simulations and their results in
\S\ref{section:numerical_solutions}.
Finally, we discuss the results and their implications in
\S\ref{section:discussion} and conclude the paper in \S\ref{section:conclusions}.

@@

\begin{equation}
\label{eq:k_short_lhs}
\frac{\widetilde{k}}{k} = 1 + \varepsilon_L \cos{\psi}
\end{equation}
Coming from the conservation of crests, as opposed to the velocity potential
superposition, the solution by \citet{longuet1960changes} requires two
approximations:
First, evaluating the long-wave velocity at $z = 0$ rather than $z = \eta_L$; and
second, further truncating the Taylor expansion series beyond the first-order
term.
These approximations underestimate the short-wave modulation magnitude
(Fig. \ref{fig:analytical_solutions_ak0.1}).
At the wave crest, (\ref{eq:k_short_exact}) yields the wavenumber modulation
that is 16.9\%, 38.5\%, 66.4\%, and 104\% larger than that of
\citet{longuet1960changes} for $\varepsilon_L = 0.1, 0.2, 0.3, 0.4$, respectively.
Although not immediately obvious from the expressions, the modulated wavenumber
(\ref{eq:k_short_lhs}) is conserved across the long-wave phase, whereas
the higher-order solution (\ref{eq:k_short_exact}) is not.
This is because combining the orbital velocities evaluated at $z=\eta_L$ and the
linearised conservation of wave crests in (\ref{eq:wavenumber}) creates excess
short-wave wavenumber.
The solution of (\ref{eq:wavenumber}) must be conservative across the long-wave
phase, as we will see later from the numerical simulations.
The steady analytical solutions for modulation of short-wave effective gravity,
amplitude, steepness, intrinsic frequency, and phase speed are also shown in
Fig. \ref{fig:analytical_solutions_ak0.1}, for reference, however, their
formulations appear in the following subsections.
Although the analytical solutions by \citet{longuet1960changes} have been updated
by the numerical results of \citet{longuet1987propagation}, they are here
nevertheless relevant to use as reference for the proposed analytical solutions.
This comparison allows quantifying solely the effect of evaluating the
long-wave properties at the surface rather than at the mean water level.

@@

\begin{equation}
\label{eq:gravity_modulation_lagrangian}
\widetilde{g_l} =
g + \frac{dW}{dt} =
g + \frac{\partial W}{\partial t} + \left(C_g + U\right) \frac{\partial W}{\partial x} =
\widetilde{g_e} + \left(C_g + U\right) \frac{\partial W}{\partial x}
\end{equation}
where $l$ and $e$ subscripts serve to denote "Lagrangian" and "Eulerian",
respectively.
Although \citet{longuet1986eulerian} thoroughly described the differences
between the Eulerian and Lagrangian effective gravities, he did so for
a fully nonlinear irrotational gravity wave, which can be computed numerically
but not analytically.
Note also that when evaluating the Lagrangian effective gravity,
\citet{longuet1987propagation} considered the orbital motion of the long wave
but not the propagation speed of the short-wave group, which can exceed the
ambient velocity $U$ depending on the properties of the short and long waves
considered.
Group speed of the short waves aside, the Lagrangian correction to the gravity
being an advective term is $\mathcal{O}(\varepsilon_L^2)$.
Third, the projection of the long-wave induced horizontal acceleration
contributes small but non-negligible effects on the short-wave effective
gravity in the curvilinear coordinate system
\citep{phillips1981dispersion,zhang1990evolution}.
These three effects are completely described by projecting the gravitational
acceleration vector from the coordinate that is perpendicular to the mean water
surface ($z=0$) to that which is perpendicular to the long-wave surface
($z=\eta_L$), and likewise for the long-wave orbital accelerations:

@@

\begin{equation}
\label{eq:gravity_modulation_linear}
\frac{\widetilde{g_e}}{g}
  = 
  \frac{
    1 - \varepsilon_L \cos{\psi} e^{\varepsilon_L \cos{\psi}}
    \left[ 1 + \left(\varepsilon_L \sin{\psi}\right)^2 e^{\varepsilon_L \cos{\psi}} \right]
  }
  {\sqrt{\left(\varepsilon_L \sin{\psi}\right)^2 + 1}}
\end{equation}
The same derivation can be carried out for the Lagrangian gravity
modulation following (\ref{eq:gravity_modulation_lagrangian})-(\ref{eq:local_slope}),
and is omitted here for brevity.
Curvilinear effects on effective gravity can be removed altogether by setting
$\alpha = 0$.
In that case, (\ref{eq:gravity_modulation_general}) simplifies to:

\begin{equation}
\label{eq:gravity_modulation_linear_no_curvature}
\frac{\widetilde{g_e}}{g} = 1 - \varepsilon_L \cos{\psi} e^{\varepsilon_L \cos{\psi}}
\end{equation}
If we evaluate the orbital accelerations at the mean water level $z=0$,
this further simplifies to:

\begin{equation}
\label{eq:gravity_modulation_linear_no_curvature_mean_level}
\frac{\widetilde{g_e}}{g} = 1 - \varepsilon_L \cos{\psi}
\end{equation}
which is the form used by \citet{longuet1960changes} and
\citet{peureux2021unsteady}, for example.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/fig_effective_gravities.pdf}
\caption{
  Analytical solutions for the effective gravitational acceleration
  modulation by long waves, as function of the long-wave phase, for $\varepsilon_L = 0.3$.
  Black is for Eulerian gravity of a linear wave evaluated at $z=0$;
  blue is the same as black but evaluated at $z=\eta$;
  orange is the same as blue but for Lagrangian gravity;
  green is the same as orange but in a curvilinear reference frame;
  red is the same as green but for a third-order Stokes wave;
  and purple is the same as red but for a fully-nonlinear wave.
  The elevation and surface velocities for the fully-nonlinear wave are computed
  following \citet{clamond2018accurate}.
  Long-wave crest and trough are located at $\psi = 0$ and $\psi = \pi$,
  respectively.
}
\label{fig:effective_gravities}
\end{figure}

@@

To quantify the contribution of the nonlinear terms in (\ref{eq:wavenumber})
and (\ref{eq:wave_action2}), and to evaluate the steadiness assumption of the
analytical solutions, we now proceed to numerically integrate the full set of
crest and action conservation equations.
(\ref{eq:wavenumber}) and (\ref{eq:wave_action2}) are discretised using
second-order central finite difference in space and integrated in time using the
fourth-order Runge-Kutta method \citep{butcher1996runge}.
The space is divided into 128 grid points.
This is effectively the same equation set and numerical configuration as that of
\citet{peureux2021unsteady}, except that their long-wave orbital velocities are
evaluated at $z=0$ instead of $z=\eta_L$, thus neglecting the Stokes drift
induced by long waves on short-wave groups
\citep{stokes1847,van2018stokes,monismith2020stokes}.
Their gravity modulation also does not consider the nonlinear or the curvature
effects, however, this difference does not qualitatively affect the results.
Another difference is the choice of the numerical scheme for spatial differences,
which in their case is the more sophisticated MUSCL4 scheme \citep{kurganov2000new}.
Although the second-order central finite difference is not appropriate for many
numerical problems, in our case it is sufficient because it is conservative and
the fields that we compute derivatives of are smooth and continuous.
The maximum relative error of the centred finite difference in this model
is $\approx 0.066\%$.
The wavenumber and wave action are conservative over time within
$\mathcal{O}(10^{-7})$ and $\mathcal{O}(10^{-5})$ relative error, respectively.
We show in the next section that our numerical solutions for infinite long-wave
trains are qualitatively equivalent to those of \citet{peureux2021unsteady}.

The numerical equations here are integrated in a fixed reference frame with
periodic boundary conditions, rather than that moving with the long-wave phase speed.
Either approach produces equivalent modulation results, however a fixed
reference frame may allow for a more intuitive interpretation of the results.
This difference is important to keep in mind when comparing the results
of \citet{peureux2021unsteady} and those presented here; in their Fig. 1 the
long-wave phase is fixed and the short waves are moving leftward
with the speed of $C_{pL} - C_g - U$ (neglecting group speed inhomogeneity),
whereas in the figures that follow, the long wave is moving rightward with its
phase speed $C_{pL}$ and the short waves (that is, their action) are moving
rightward with the speed of $C_g + U$ (neglecting group speed inhomogeneity).
The crest and action conservation equations are integrated in the curvilinear
coordinate system in which the horizontal coordinate follows the long-wave
surface, akin to \citet{zhang1990evolution}.

@@

The first set of simulations is performed for the long-wave steepness of
$\varepsilon_L = 0.1$, $0.2$, and $0.4$ (Fig. \ref{fig:numerical_solutions}).
The long-waves are initialised from rest ($a_L = 0$) and gradually ramped up
to their target steepness over 5 long-wave periods, thus emulating the passage
of a long-wave group
(the importance of which we discuss in more detail in the next subsection).
At $\varepsilon_L = 0.1$, the numerical solutions for the wavenumber, amplitude,
and steepness modulation are all remarkably similar to the analytical solutions
(Fig. \ref{fig:numerical_solutions}a,d,g).
This suggests that for low long-wave steepness the steady analytical solution
is a reasonable approximation for the fully-nonlinear crest-action conservation
equation solutions.
The numerical solutions diverge more notably from the analytical ones at
moderate steepness of 0.2 (Fig. \ref{fig:numerical_solutions}b,e,h), and are
considerably different at very high steepness of 0.4
(Fig. \ref{fig:numerical_solutions}c,f,i).
These differences suggest that at high steepness the inhomogeneity of
gravitational acceleration and group speed, otherwise ignored in the steady
solutions, become important.

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/numerical_vs_analytical_solutions_k.pdf}
\caption{
  Maximum wavenumber modulation as function of long-wave steepness $\varepsilon_L$.
  Black line and circles are for the analytical and numerical solutions from
  \citet{longuet1960changes} and \citet{longuet1987propagation}, respectively.
  Blue line is based on the analytical solutions from this paper.
  Green and orange lines are for the numerical solutions from this paper
  using linear and third-order Stokes approximations of long waves, respectively.
  Red line is for the fully-nonlinear gravity wave based on \citet{clamond2018accurate}.
}
\label{fig:numerical_vs_analytical_k}
\end{figure}

@@

The maximum modulations of the short-wave wavenumber, amplitude, and steepness
from the analytical and numerical solutions, are shown on Figs.
\ref{fig:numerical_vs_analytical_k},
\ref{fig:numerical_vs_analytical_a}, and
\ref{fig:numerical_vs_analytical_ak}, respectively, as function of the long-wave
steepness $\varepsilon_L$.
Overall, the numerical solutions, based on the linear and the third-order Stokes
long waves alike, begin to diverge from the analytical solution at
$\varepsilon_L \approx 0.15$.
The numerical solution using the third-order Stokes long wave is quantitatively
very similar to the fully-nonlinear solutions of \citet{longuet1987propagation}
in wavenumber modulation, however with somewhat lower amplitude and steepness
modulations.
Using the velocity and gravity properties of a fully-nonlinear wave, based on
the approach by \citet{clamond2018accurate}, the numerical solutions diverge
from those of the third-order Stokes wave only at very large steepnesses
($\varepsilon_L > 0.3$).
Using a fully-nonlinear form of the long-wave thus does not collapse the
difference between the solutions here and those of \citet{longuet1987propagation}.
For example, at $\varepsilon_L = 0.4$, the steepness modulation in our solution
is 6.6, whereas \citet{longuet1987propagation} finds it to be 7.8.
A possible reason for this difference is that the solutions by
\citet{longuet1987propagation} are stationary (see his \S2), whereas in our
numerical model we solve the prognostic nonlinear equations for the short-wave
wavenumber and action.
As we will see in the next section, and as was suggested by \citet{peureux2021unsteady},
they are often, but not always, mostly steady, and can be strongly unsteady in
special cases.
Nevertheless, the overall steepness modulation between the two numerical
approaches are remarkably similar, and either solution would cause the
short-waves of even low steepness to break \citep{banner1993wave}.
Where the two modulation solutions differ (at $\varepsilon_L > 0.2$),
either solution would lead to short-wave breaking after one long-wave period.
This suggests that in the cases where the short-waves would survive the
modulation and persist over multiple long-wave periods, the two solutions are
effectively equivalent.

As the numerical solutions are for the nonlinear wave action and crest conservation
equations, it is important to re-evaluate the homogeneity and stationarity
requirements for the wave action conservation to remain valid.
Fig. \ref{fig:homogeneity_stationarity_numerical} shows the homogeneities
and stationarities of the short-wave wavenumber, action, and effective gravity
as function of the long-wave steepness $\varepsilon_L$ and the wavenumber ratio
$k/k_L$, based on the numerical simulations.
Although homogeneity is slightly lower ($H_{k,N,g} \approx 0.92$ for $k/k_L = 10$
and $\varepsilon_L = 0.4$) compared to the analytical solutions,
the stationarity is significantly reduced.
Following the same criteria for strong ($S_{k,N,g} > 0.99$) and weak
($S_{k,N,g} > 0.9$) stationarity as in \S\ref{sec:wave_action_conservation},
the numerical solutions are only weakly stationary for $\varepsilon_L < 0.21$
at $k/k_L = 10$, and for $\varepsilon_L < 0.36$ at $k/k_L = 100$.
This provides a quantitative guidance on when the short-wave action can be
considered to be conserved in the presence of longer waves.
